---
title: "placentaImmuneCellRefEPIC450k Vignette"
author: "Kyle Abraham Campbell"
date: "2024-12-03"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{placentaImmuneCellRefEPIC450k Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

placentaImmuneCellRefEPIC450k contains a Illumina DNA methylation microarray-based
reference meant for use in estimating the cell composition of whole placenta
villous tissue samples in epidemiological studies.

This package is a companion package to "Placental and Immune Cell DNA
Methylation Reference Panel for Bulk Tissue Cell Composition Estimation in
Epidemiological Studies", DOI:http://dx.doi.org/10.1080/15592294.2024.2437275.

Upon loading the library, a DNA methylation signature matrix called "ref" will be loaded into memory,
which contains a 721 row (each row represents a common CpG from the Illumina
450k and EPIC DNA methylation microarrays) by 12 column (each column represents
a placenta or maternal immune cell type present at the maternal-fetal interface)
matrix of beta values expected for each cell type at that CpG.

This reference includes 5 placental cell types (cytotrophoblast, endothelial, Hofbauer, stromal, and syncytiotrophoblast), 6 peripheral immune cell types (B cell, CD4T, CD8T, monocyte, neutrophil, and natural killer), and nucleated red blood cells. 

## Application
This reference is built for plug and play use with the EpiDISH deconvolution package, but can be used with any deconvolution algorithm.
```{r install_packages, eval = F}
BiocManager::install("placentaImmuneCellRefEPIC450k")
BiocManager::install("EpiDISH")
library(EpiDISH)
```

```{r load_packages}
library(placentaImmuneCellRefEPIC450k)
```

Simply by loading the placentaImmuneCellRefEPIC450k package, you will be ready to deconvolute your bulk tissue beta matrix with EpiDISH.
```{r, eval = F}
estimates <- epidish(beta.m = bulk,
                     ref.m = ref)
```

For advanced users who would like to create their own reference using our data, the full cell type-specific data and metadata can be manually downloaded from GEO Accession number GSE271697. Specifically, GSE271697_beta.tsv.gz_beta.tsv.gz. Unfortuantely, due to file size limitations, we cannot host this file directly on GitHub.

We still provide the phenotype metadata readily accessible here.
```{r}
data(pd, package = "placentaImmuneCellRefEPIC450k") # Metadata
head(p.pd)
```

Session info:
```{r}
sessionInfo()
```

